---
title: "01 — Xenium Spatial Transcriptomics"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    self-contained: true
    title-block-banner: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

# Context

- **Scope**: standalone Xenium FFPE Human Breast Cancer Rep1 processing — from raw 10X outs through to a clustered Giotto object carrying the post-Xenium H&E and 3-channel IF images
- **Upstream**: dataset pulled from `https://cf.10xgenomics.com/samples/xenium/1.0.1/Xenium_FFPE_Human_Breast_Cancer_Rep1/Xenium_FFPE_Human_Breast_Cancer_Rep1_outs.zip`
- **Reference walkthrough**: adapted from [giottosuite.com/articles/xenium_breast_cancer](https://giottosuite.com/articles/xenium_breast_cancer.html) — parameters and ordering follow the Giotto tutorial, with local modifications for image alignment and group conventions
- **Downstream**: the saved Giotto object (`../checkpoints/01-xenium-gobj/`) feeds script `04-integration.qmd` for the joint harmony-integrated workflow

## Conventions

- **Checkpoint storage**: `saveGiotto()` / `loadGiotto()` are used throughout for Giotto objects
  - `write_rds()` / `read_rds()` are unsafe here: Giotto objects contain pointer-based assets (`terra` raster images, arrow-backed transcripts) that do not re-hydrate correctly through `saveRDS`
  - `saveGiotto()` writes a self-contained folder with each component serialised separately; `loadGiotto(..., reconnect_giottoImage = TRUE)` restores the image pointers against the on-disk TIFFs

| Stage | Path | Method |
|---|---|---|
| Xenium processed (this script output) | `../checkpoints/01-xenium-gobj/` | `saveGiotto` |
| scRNA-Seq Seurat (script 02 output) | `../checkpoints/02-sc-seu-annotated.rds` | `write_rds` |
| scRNA-Seq Giotto annotated (script 03 output) | `../checkpoints/03-sc-giotto-annotated/` | `saveGiotto` |
| Joined + Harmony + label transfer (script 04 output) | `../checkpoints/04-joined-giotto-harmony-annotated/` | `saveGiotto` |

- **Figure output**: all plots go to `../write/figures/` with explicit `save_name`, `base_width`, and `base_height` set via `save_param`
- **Figure prefix**: `01-xenium-*` for every figure produced here; downstream scripts use `02-`, `03-`, `04-`

# Setup

## Arrow Installation

- **Purpose**: Giotto reads Xenium `transcripts.parquet` through `arrow`, which must have ZSTD compression support — 10X parquet files are ZSTD-compressed
- **Check logic**: if `arrow` is missing or was built without ZSTD, reinstall from the Apache R-universe with `ARROW_WITH_ZSTD=ON`
- **One-off**: this usually runs once per machine; after a successful build the check short-circuits

```{r}
#| label: arrow-install

# Libraries ----------
# (none needed — this chunk stays self-contained on base R)

# Processing ----------
has_arrow <- requireNamespace("arrow", quietly = TRUE)
zstd <- TRUE
if (has_arrow) {
    zstd <- arrow::arrow_info()$capabilities[["zstd"]]
}
if (!has_arrow || !zstd) {
    Sys.setenv(ARROW_WITH_ZSTD = "ON")
    install.packages(
        "arrow",
        repos = c("https://apache.r-universe.dev"),
        type = "source"
    )
}
```

## Python Environment and Giotto Instructions

- **Python backend**: Giotto delegates image and IO work to Python via `reticulate` — `tifffile` handles OME-TIFF conversion, `imagecodecs` provides JPEG2000 decoding, `scrublet` / `leidenalg` / `python-igraph` cover dimensionality-reduction and clustering backends
- **Env management**: this project uses `uv` (see `pyproject.toml`), not conda — `installGiottoEnvironment()` is intentionally not called because it would create a parallel conda env
- **Pointing Giotto at the uv env**: `set_giotto_python_path(".venv/bin/python")` — must run before any Giotto operation that dispatches to Python
- **Instructions object (`instrs`)**: centralises the save behaviour (plot directory, save/show/return flags, python path) so every Giotto plotting call inherits the same defaults

```{r}
#| label: setup-giotto

# Libraries ----------
library(Giotto)

# Processing ----------
python_path <- ".venv/bin/python"
set_giotto_python_path(python_path = python_path)

results_folder <- "../write/figures/"

instrs <- createGiottoInstructions(
    save_dir = results_folder,
    save_plot = TRUE,
    show_plot = FALSE,
    return_plot = FALSE,
    python_path = python_path
)
```

# Load Xenium Data

- **Input**: the full unzipped Xenium bundle at `../read/outs/` — includes `cell_feature_matrix.h5`, `transcripts.parquet`, polygon parquet, and the `morphology*.ome.tif` stack
- **`qv_threshold = 20`**: 10X's own default; filters transcripts whose per-read Phred-scaled quality value is below 20 (i.e. <99% confidence in decoding)
- **Feature types and split keywords**: Xenium panels ship four feature categories; keeping them separate lets us inspect each in isolation during QC
  - `rna` — real gene probes
  - `UnassignedCodeword` — decoded codewords not assigned to a target (noise floor)
  - `NegControlCodeword` — codewords without matching probes (decoding error rate)
  - `NegControlProbe` + `antisense` — probes without target sequences (non-specific binding rate)
- **Repo structure caveat**: `xeniumranger` versions newer than 1.0.1 reshuffle the `outs/` tree; adjust `xenium_dir` accordingly if upgrading

```{r}
#| label: load-xenium

# Libraries ----------
library(Giotto)

# Inputs ----------
data_path <- "../read/"

# Processing ----------
# Feature-type splits are specific to pre-release / early Xenium panels;
# the 4 codeword categories isolate noise from true RNA signal.
feat_types <- c(
    "rna",
    "UnassignedCodeword",
    "NegControlCodeword",
    "NegControlProbe"
)
split_keywords <- list(
    c("BLANK"),
    c("NegControlCodeword"),
    c("NegControlProbe", "antisense")
)

xenium_gobj <- createGiottoXeniumObject(
    xenium_dir = file.path(data_path, "outs"),
    qv_threshold = 20,
    feat_type = feat_types,
    split_keyword = split_keywords,
    # Aligned images are attached in a later chunk (see "Post-Xenium Aligned Images")
    # rather than passed here, to keep the conversion step visible and auditable.
    instructions = instrs
)

force(xenium_gobj)
```

# Initial Visualisation

- **Overview plot**: a flat scatter of all cell centroids — confirms the tissue footprint loaded correctly and the coordinate frame is sane (Xenium uses a negative-y convention)
- **Zoomed region**: a 1000×1000 µm window with polygon overlays on the morphology image — confirms cell segmentation boundaries align with the DAPI channel
  - `xlim = c(1000, 2000)` / `ylim = c(-3000, -2000)` was chosen to land on a region with mixed tumour/stromal morphology for visual QC

```{r}
#| label: initial-viz

# Libraries ----------
library(Giotto)

# Processing ----------
# Full-field centroid overview — no polygons, no image; fastest sanity check.
spatPlot2D(
    xenium_gobj,
    point_shape = "no_border",
    point_size = 1,
    save_param = list(
        save_name = "01-xenium-overview-spatial",
        base_width = 10,
        base_height = 8
    )
)

# Zoomed view with polygon outlines on the morphology image — checks segmentation
# quality against the underlying DAPI signal.
spatInSituPlotPoints(
    xenium_gobj,
    show_image = TRUE,
    polygon_line_size = 0.1,
    polygon_color = "#BB0000",
    polygon_alpha = 0.2,
    xlim = c(1000, 2000),
    ylim = c(-3000, -2000),
    save_param = list(
        save_name = "01-xenium-zoomed-polygons-overview",
        base_width = 8,
        base_height = 8
    )
)
```

# Post-Xenium Aligned Images

- **Group motivation**: our in-house liver workflow produces high-quality post-Xenium segmentation via `cellpose3`; testing that those polygons can drive the Giotto count matrix is an open question this project is scaffolding toward
- **This dataset**: the 10X breast-cancer sample ships post-Xenium H&E and a 3-channel IF (DAPI / HER2 / CD20) with per-image alignment CSVs produced by Xenium Explorer
- **Why convert OME-TIFF → TIFF**: Giotto reads flat `.tif` natively via `terra`; `.ome.tif` metadata has to be parsed through Python `tifffile`, which introduces a noticeable speed penalty at every access — converting up front trades one cost for many

## Converting OME-TIFF

- **`GiottoClass::ometif_to_tif()`**: uses Python `tifffile` / `imagecodecs` as the backend; writes flat TIFFs to `read/tif_exports/` by default
- **Multi-page IF**: the IF OME-TIFF is a 3-page stack (one page per channel) — each page is converted separately with `page = p`
- **`ometif_metadata(..., node = "Channel")$Name`**: pulls the channel names (DAPI, HER2, CD20) from the OME-TIFF XML so we can name the resulting Giotto images correctly

```{r}
#| label: convert-ome-tiff

# Libraries ----------
library(Giotto)
library(purrr)

# Inputs ----------
path_he <- "../read/Xenium_FFPE_Human_Breast_Cancer_Rep1_he_image.ome.tif"
path_if <- "../read/Xenium_FFPE_Human_Breast_Cancer_Rep1_if_image.ome.tif"

# Processing ----------
# H&E — single-page; converts to read/tif_exports/<basename>.tif by default.
conv_path_he <- GiottoClass::ometif_to_tif(path_he)

# IF — 3-page stack (DAPI / HER2 / CD20); convert each page separately.
conv_path_if <- map(1:3, function(p) {
    GiottoClass::ometif_to_tif(
        path_if,
        page = p,
        overwrite = TRUE
    )
})

# Pull channel names from the OME-TIFF XML metadata so the attached Giotto images
# inherit biologically meaningful names rather than "page_1", "page_2", etc.
if_channels <- GiottoClass::ometif_metadata(path_if, node = "Channel")$Name
```

## Loading Aligned Images

- **`importXenium()`**: Giotto helper that returns a list of loaders tied to an outs directory
- **`load_aligned_image()`**: reads the flat TIFF and applies the affine transform from the per-image `imagealignment.csv` (produced by Xenium Explorer) to place the image in the Xenium coordinate frame
- **Per-channel treatment**: each IF channel is loaded as a separate Giotto image named after the stain (DAPI, HER2, CD20), enabling channel-specific overlay in later plots

```{r}
#| label: load-aligned-images

# Libraries ----------
library(Giotto)

# Inputs ----------
data_path <- "../read/"

# Processing ----------
x <- importXenium(file.path(data_path, "outs"))

# H&E — one image, one alignment file.
img_he <- x$load_aligned_image(
    name = "post_he",
    path = conv_path_he,
    imagealignment_path = "../read/Xenium_FFPE_Human_Breast_Cancer_Rep1_he_imagealignment.csv"
)

# IF — iterate over the 3 channels; all share the same alignment CSV.
img_if <- lapply(1:3, function(if_i) {
    x$load_aligned_image(
        name = if_channels[[if_i]],
        path = conv_path_if[[if_i]],
        imagealignment_path = "../read/Xenium_FFPE_Human_Breast_Cancer_Rep1_if_imagealignment.csv"
    )
})
```

## Appending Images to the Giotto Object

- **Cells and transcripts still come from Xenium**: attaching these images does not alter the segmentation or count matrix; they are overlay assets for visualisation and for any future workflow that uses them to drive segmentation
- **Open question**: whether polygons produced from an external image (e.g. `cellpose3` on the H&E) can replace the Xenium-provided `cell_boundaries` and then be used via `calculateOverlapRaster` to build an alternative count matrix — not validated here, flagged for follow-up

```{r}
#| label: attach-images

# Libraries ----------
library(Giotto)

# Processing ----------
xenium_gobj <- setGiotto(
    xenium_gobj,
    c(
        list(img_he),
        img_if
    )
)

# Confirm the image slots were populated with the expected names.
print(names(xenium_gobj@images))

# Visual check: HER2-overlaid polygons in the same zoomed region used earlier.
# HER2 is a clinically actionable receptor (amplified in ~15-20% of breast cancers);
# its IF signal should localise to epithelial/cancer cell polygons.
spatInSituPlotPoints(
    xenium_gobj,
    xlim = c(1000, 2000),
    ylim = c(-3000, -2000),
    show_image = TRUE,
    image_name = "HER2",
    polygon_line_size = 0.1,
    polygon_color = "#BB0000",
    polygon_alpha = 0.2,
    save_param = list(
        save_name = "01-xenium-zoomed-polygons-her2",
        base_width = 8,
        base_height = 8
    )
)

# DAPI overlay for the same region — nuclear counterstain, should fill every polygon.
spatInSituPlotPoints(
    xenium_gobj,
    xlim = c(1000, 2000),
    ylim = c(-3000, -2000),
    show_image = TRUE,
    image_name = "dapi",
    polygon_line_size = 0.1,
    polygon_color = "blue",
    polygon_alpha = 0.2,
    save_param = list(
        save_name = "01-xenium-zoomed-polygons-dapi",
        base_width = 8,
        base_height = 8
    )
)
```

# Data Aggregation

- **Step**: convert the point-cloud of decoded transcripts + polygon boundaries into a cell × gene count matrix
- **`calculateOverlapRaster`**: intersects `rna` transcript points with `cell` polygons using a raster-based algorithm — faster than vector-polygon intersection for dense point clouds
- **`overlapToMatrix`**: collapses the intersection records into a sparse count matrix stored under `expression$raw`; this matrix is what downstream filtration, normalisation, and clustering operate on

```{r}
#| label: data-aggregation

# Libraries ----------
library(Giotto)

# Processing ----------
# Raster-based overlap: bin transcripts into the raster, then assign to polygons.
xenium_gobj <- calculateOverlapRaster(
    gobject = xenium_gobj,
    spatial_info = "cell",
    feat_info = "rna"
)

# Collapse overlap records into a sparse cell × gene count matrix.
xenium_gobj <- overlapToMatrix(
    xenium_gobj,
    poly_info = "cell",
    feat_info = "rna",
    name = "raw"
)

showGiottoExpression(xenium_gobj)
```

# Filtration, Normalisation, and Dimensionality Reduction

- **Filtration thresholds** (follow the Giotto tutorial defaults):
  - `expression_threshold = 1`: a feature counts as detected if ≥1 transcript is assigned to the cell
  - `feat_det_in_min_cells = 3`: drop genes detected in fewer than 3 cells — removes genes with no population-level signal
  - `min_det_feats_per_cell = 5`: drop cells expressing fewer than 5 genes — removes empty or blank polygons
- **Normalisation**: Giotto's default is library-size normalisation + log-transform + per-cell scaling — adequate for Xenium's ~300-gene panel where SCT-style variance stabilisation is overkill
- **`addStatistics()`**: computes per-cell and per-gene summary stats needed by downstream plotting
- **PCA on all features (`feats_to_use = NULL`)**: targeted panels are small enough to PCA without HVF selection; using all genes avoids arbitrarily throwing away panel content

```{r}
#| label: filter-normalise-pca

# Libraries ----------
library(Giotto)

# Processing ----------
xenium_gobj <- xenium_gobj |>
    filterGiotto(
        spat_unit = "cell",
        expression_threshold = 1,
        feat_det_in_min_cells = 3,
        min_det_feats_per_cell = 5
    ) |>
    normalizeGiotto() |>
    addStatistics() |>
    runPCA(feats_to_use = NULL)

# Scree plot — inspect the PCA eigenvalue elbow to pick `dimensions_to_use`.
screePlot(
    xenium_gobj,
    ncp = 20,
    save_param = list(
        save_name = "01-xenium-scree-plot",
        base_width = 7,
        base_height = 5
    )
)

# First two PCs — quick visual check that PCA captured population structure.
plotPCA(
    xenium_gobj,
    point_size = 0.1,
    save_param = list(
        save_name = "01-xenium-pca",
        base_width = 7,
        base_height = 6
    )
)
```

| Parameter | Value | Justification |
|---|---|---|
| `expression_threshold` | 1 | Single transcript is sufficient evidence for detection at Xenium sensitivity |
| `feat_det_in_min_cells` | 3 | Removes genes with no population-level signal without dropping rare markers |
| `min_det_feats_per_cell` | 5 | Removes empty polygons / decoding failures |
| `feats_to_use` (PCA) | NULL (all genes) | Targeted panel of ~300 genes — no need for HVF pre-selection |

## Further Dimensional Reduction

- **`dimensions_to_use = 1:10`**: a conservative default that works well for the breast cancer panel; the scree plot above should be consulted before changing it
- **Running both tSNE and UMAP**: UMAP preserves global structure better, tSNE preserves local neighbourhoods — looking at both is a cheap sanity check for embedding stability

```{r}
#| label: umap-tsne

# Libraries ----------
library(Giotto)

# Processing ----------
xenium_gobj <- runtSNE(xenium_gobj, dimensions_to_use = 1:10)
xenium_gobj <- runUMAP(xenium_gobj, dimensions_to_use = 1:10)

plotTSNE(
    xenium_gobj,
    point_size = 0.01,
    save_param = list(
        save_name = "01-xenium-tsne",
        base_width = 7,
        base_height = 6
    )
)

plotUMAP(
    xenium_gobj,
    point_size = 0.01,
    save_param = list(
        save_name = "01-xenium-umap",
        base_width = 7,
        base_height = 6
    )
)
```

# Clustering

- **Graph-based clustering**: Giotto follows the standard scRNA-Seq recipe — build a kNN graph on the PCA embedding, then run Leiden on the graph
- **`k = 10`**: neighbourhood size for the kNN graph; small enough to preserve fine structure, large enough to connect most cells
- **`resolution = 0.25`**: chosen to yield ~8–12 coarse populations — suits the 298-gene panel, which cannot resolve fine subtypes regardless
- **`n_iterations = 100`**: Leiden optimisation iterations; default in Giotto, robust to the parameter

```{r}
#| label: leiden-clustering

# Libraries ----------
library(Giotto)

# Processing ----------
xenium_gobj <- xenium_gobj |>
    createNearestNetwork(
        dimensions_to_use = 1:10,
        k = 10
    ) |>
    doLeidenCluster(
        resolution = 0.25,
        n_iterations = 100
    )

plotTSNE(
    xenium_gobj,
    cell_color = "leiden_clus",
    show_legend = FALSE,
    point_size = 0.01,
    point_shape = "no_border",
    save_param = list(
        save_name = "01-xenium-tsne-leiden-clusters",
        base_width = 7,
        base_height = 6
    )
)

plotUMAP(
    xenium_gobj,
    cell_color = "leiden_clus",
    show_legend = FALSE,
    point_size = 0.01,
    point_shape = "no_border",
    save_param = list(
        save_name = "01-xenium-umap-leiden-clusters",
        base_width = 7,
        base_height = 6
    )
)
```

# Spatial Visualisation with Clusters

- **Centroid plot with DAPI backdrop**: Leiden clusters coloured onto their spatial positions with the DAPI IF image as the background — reveals whether clusters correspond to tissue-level spatial domains or mix uniformly
- **`plot_method = "scattermore"`**: GPU-accelerated scatter backend; essential for rendering hundreds of thousands of points without choking
- **Polygon plot**: the same clustering rendered as filled cell polygons instead of centroids — preserves cell morphology at the cost of being much slower to render

```{r}
#| label: spatial-viz-clusters

# Libraries ----------
library(Giotto)

# Processing ----------
spatPlot2D(
    xenium_gobj,
    plot_method = "scattermore",
    cell_color = "leiden_clus",
    point_size = 0.5,
    point_shape = "no_border",
    image = "DAPI",
    background_color = "black",
    save_param = list(
        save_name = "01-xenium-spatial-leiden-clusters",
        base_width = 10,
        base_height = 8
    )
)

# Polygon rendering — biologically more informative (cell shape is preserved),
# but noticeably slower than centroid scattering for dense fields.
spatInSituPlotPoints(
    xenium_gobj,
    polygon_feat_type = "cell",
    polygon_alpha = 1,
    polygon_line_size = 0.01,
    polygon_color = "black",
    polygon_fill = "leiden_clus",
    polygon_fill_as_factor = TRUE,
    save_param = list(
        save_name = "01-xenium-polygons-leiden-clusters",
        base_width = 10,
        base_height = 10
    )
)
```

# Marker Gene Visualisation

- **Purpose**: confirm that the clustering has captured the expected breast-tumour-microenvironment populations before committing to any annotation
- **Panel**: curated lineage markers covering the populations expected in FFPE breast cancer — lymphoid, myeloid, stromal, vascular, and epithelial/cancer compartments
- **Marker choices** (grouped by compartment):
  - **T cells**: `CD3E` (pan-T), `CD8A` (cytotoxic), `CD4` (helper)
  - **B lineage**: `MS4A1` (CD20, mature B), `JCHAIN` (plasmablast / antibody-secreting cell)
  - **Cancer-associated fibroblasts**: `DCN`, `SFRP2` (mesenchymal CAFs), `TAGLN` (contractile CAF / perivascular-like)
  - **Endothelium**: `RAMP2` (blood endothelial), `PROX1` (lymphatic endothelial)
  - **Myeloid**: `TYROBP` (DAP12 adaptor, pan-myeloid)
  - **Mast cells**: `CPA3`, `TPSAB1` (granule proteases)
  - **Epithelial / cancer**: `KRT8` (luminal keratin), `MKI67` (proliferation), `SPINT2` (epithelial serine-protease inhibitor), `CD24` (also a breast cancer stemness marker)
  - **Myoepithelial / basal**: `SOX10`, `TP63`
  - **Checkpoint**: `PDCD1` (PD-1 — immune exhaustion / ICI target)

```{r}
#| label: marker-genes

# Libraries ----------
library(Giotto)

# Processing ----------
dimFeatPlot2D(
    xenium_gobj,
    feats = c(
        "CD3E", "CD8A", "CD4",        # T cell (pan, cytotoxic, helper)
        "JCHAIN",                      # plasmablast
        "MS4A1",                       # B (CD20)
        "DCN", "SFRP2",                # CAF (mesenchymal)
        "TAGLN",                       # CAF / PVL (contractile)
        "RAMP2",                       # blood endothelial
        "PROX1",                       # lymphatic endothelial
        "TYROBP",                      # myeloid (DAP12)
        "CPA3", "TPSAB1",              # mast cell
        "SPINT2", "MKI67", "KRT8",     # epithelial / cancer (proliferation + keratin)
        "SOX10", "TP63",               # myoepithelial / basal
        "CD24",                        # B cell + cancer stemness
        "PDCD1"                        # PD-1 (T cell exhaustion / ICI target)
    ),
    point_size = 0.2,
    point_border_stroke = 0,
    background_color = "black",
    cow_n_col = 4,
    show_legend = FALSE,
    gradient_style = "sequential",
    save_param = list(
        save_name = "01-xenium-marker-genes",
        base_width = 14,
        base_height = 16
    )
)
```

# Save Checkpoint

- **`saveGiotto()` over `write_rds()`**: mandatory for Giotto objects — the attached post-Xenium H&E and IF images are `terra` raster pointers, and arrow-backed transcript tables are also pointer-based; `write_rds` silently produces a reloadable-but-broken object
- **Folder layout**: `saveGiotto` writes a directory (`foldername`) containing one serialised component per Giotto slot plus an image subfolder with the re-connectable TIFF paths
- **Reload**: downstream scripts use `loadGiotto("../checkpoints/01-xenium-gobj", reconnect_giottoImage = TRUE, python_path = ".venv/bin/python")` to re-hydrate the object with image pointers intact
- **Caveat**: `reconnect_giottoImage = TRUE` assumes the `read/tif_exports/` TIFFs have not moved — if the project is relocated, re-run the "Converting OME-TIFF" chunk before reloading

```{r}
#| label: save-checkpoint

# Libraries ----------
library(Giotto)

# Outputs ----------
saveGiotto(
    gobject = xenium_gobj,
    dir = "../checkpoints",
    foldername = "01-xenium-gobj",
    method = "saveRDS"
)
```
