---
title: "02 — scRNA-Seq Analysis and Annotation (Seurat)"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    self-contained: true
    title-block-banner: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

# Context

- **Scope**: Seurat-based processing of the matched scRNA-Seq dataset (`GSM7782696`, 5′ 10X chemistry, same breast tumour indication) — the purpose is to produce **cluster-level cell type annotations** that will be transferred into a Giotto object downstream
- **Why Seurat, not Giotto**: two earlier paths were attempted and dropped
  - **Giotto-only clustering**: produced Leiden partitions that did not match the published annotations, leaving no clean way to align reference labels back to our clusters
  - **`GPTCelltype`**: the GPT-4 annotation of Giotto marker tables gave unstable / unspecific calls and did not improve on manual marker review
  - **`CyteTypeR`** was the intended annotation engine but requires a Seurat object as input — which is the immediate reason the scRNA-Seq preprocessing was moved over to Seurat
- **Current annotation method**: manual `case_when` mapping from clusters → cell types, driven by inspecting curated lineage markers; `CyteTypeR` can be plugged in later
- **Upstream dependency**: none — this script is independent and can be run before or after `01-xenium-analysis.qmd`
- **Downstream consumer**: `03-single-cell-giotto.qmd` loads `../checkpoints/02-sc-seu-annotated.rds` to copy its `celltype` labels onto a parallel Giotto scRNA-Seq object

## Checkpoint Convention

| Stage | File | Description |
|---|---|---|
| Annotated Seurat object (this script output) | `../checkpoints/02-sc-seu-annotated.rds` | Seurat object with `percent.mt`, clustering, and `celltype` metadata |

# Data Import

- **Input**: `../read/GSM7782696_5p_count_filtered_feature_bc_matrix.h5` — 10X filtered matrix from the matched scRNA-Seq sample
- **`Read10X_h5`** returns a sparse matrix of raw counts; `CreateSeuratObject` wraps it in the v5 layered object

```{r}
#| label: load-data

# Libraries ----------
library(Seurat)
library(SeuratObject)

# Inputs ----------
sc_seu_counts <- Read10X_h5(
    filename = "../read/GSM7782696_5p_count_filtered_feature_bc_matrix.h5"
)

# Processing ----------
sc_seu <- CreateSeuratObject(counts = sc_seu_counts)
```

# Quality Control

- **Metric used**: `percent.mt` only — `PercentageFeatureSet(pattern = "^MT-")` tags mitochondrial transcripts and computes their share of total UMI per cell
- **Why a single fixed threshold here (`< 50%`)**: the matched Xenium panel does not itself have mitochondrial probes, and the Giotto-only arm's QC was already permissive (`mito < 50%` kept 17,963 cells vs. 5,006 dropped); matching that threshold keeps the two single-cell arms comparable if both are ever revisited
- **Not used**: MAD-based filtration on `nCount_RNA` / `nFeature_RNA` — deliberately omitted to match the Giotto arm and because the downstream annotation transfer only cares about cell-level labels, not QC subtlety
- **Open**: a MAD-driven pass could be added later (see `scater::isOutlier`), but would risk diverging from the Giotto-side cell list that enters the joint object in script 04

```{r}
#| label: qc-mito

# Libraries ----------
library(Seurat)

# Processing ----------
# Mitochondrial fraction — high values indicate stressed / dying cells where
# cytoplasmic mRNA has leaked out and only mitochondrial reads remain.
sc_seu[["percent.mt"]] <- PercentageFeatureSet(sc_seu, pattern = "^MT-")

# Fixed 50% cutoff chosen to match the Giotto arm's mito filter.
sc_seu_subset <- subset(
    sc_seu,
    subset = percent.mt < 50
)
```

| Parameter | Value | Justification |
|---|---|---|
| Mitochondrial pattern | `^MT-` | Human nuclear-encoded mito prefix (GENCODE) |
| `percent.mt` cutoff | `< 50%` | Permissive; matches the Giotto arm for cross-arm comparability |

# Processing Pipeline

- **Normalisation**: `NormalizeData` (log-normalisation, default scale factor 10,000) — not SCTransform, because the simpler LogNormalize path is what the Giotto arm uses and keeps the two pipelines numerically comparable
- **`FindVariableFeatures`**: default `vst` method, 2,000 features — adequate for a homogeneous single-tissue dataset
- **`ScaleData`**: z-scores variable features for PCA input
- **`RunPCA(npcs = 20)`**: 20 PCs is sufficient for a well-characterised breast-tumour dataset; the elbow plot typically flattens before PC 15
- **`FindNeighbors(dims = 1:20)` + `RunUMAP(dims = 1:20)`**: standard graph and embedding construction on the PCA subspace

```{r}
#| label: process

# Libraries ----------
library(Seurat)

# Processing ----------
sc_seu_subset <- sc_seu_subset |>
    NormalizeData() |>
    FindVariableFeatures() |>
    ScaleData() |>
    RunPCA(npcs = 20) |>
    FindNeighbors(dims = 1:20) |>
    RunUMAP(dims = 1:20)
```

| Parameter | Value | Justification |
|---|---|---|
| Normalisation | `NormalizeData` (LogNormalize) | Matches Giotto arm; avoids SCT vs. LogNorm divergence |
| `npcs` | 20 | Elbow plateau for this panel; adding more adds noise |
| `dims` (UMAP / neighbours) | 1:20 | Full PCA subspace |

# Clustering

- **`FindClusters(resolution = 0.1)`**: low resolution deliberately — the downstream purpose is **label transfer onto Xenium cells**, which carry a ~300-gene panel unable to resolve fine subtypes
  - Giving the Seurat side finer clusters than Xenium can support would only inject noise at the transfer step
  - A resolution of 0.1 yields ~10 clusters, which aligns with the granularity Xenium can sustain

```{r}
#| label: cluster

# Libraries ----------
library(Seurat)

# Processing ----------
sc_seu_subset <- sc_seu_subset |>
    FindClusters(resolution = 0.1)

# UMAP coloured by cluster — quick check that clusters separate cleanly.
p_umap_clusters <- sc_seu_subset |>
    BadranSeq::do_DimPlot(
        reduction = "umap",
        group.by = "seurat_clusters"
    )

ggplot2::ggsave(
    filename = "../write/figures/02-sc-seu-umap-clusters.png",
    plot = p_umap_clusters,
    width = 8,
    height = 6,
    bg = "white"
)
```

# Marker Gene Visualisation

- **Purpose**: visualise the same curated panel used in script `01` on the scRNA-Seq UMAP — this is the primary input to the manual cluster-to-celltype mapping below
- **Why identical marker list to script `01`**: keeps the biology frame shared between the scRNA-Seq and Xenium arms; every marker here is also a feature on the Xenium panel
- **`BadranSeq::do_FeaturePlot`**: preferred visual for single-feature UMAP overlays per the group's style guide; falls back to Seurat's `FeaturePlot` if unavailable
- **Panel size**: 18 features × 3 columns → 6 rows; generous `height = 30` inches keeps individual panels legible

```{r}
#| label: marker-genes

# Libraries ----------
library(Seurat)
library(ggplot2)

# Processing ----------
feats <- c(
    "CD3E", "CD8A", "CD4",       # T cell (pan, cytotoxic, helper)
    "JCHAIN",                     # plasmablast
    "MS4A1",                      # B cell (CD20)
    "DCN", "SFRP2",               # CAF (mesenchymal)
    "TAGLN",                      # CAF / PVL (contractile)
    "RAMP2",                      # blood endothelial
    "PROX1",                      # lymphatic endothelial
    "TYROBP",                     # myeloid (DAP12)
    "CPA3", "TPSAB1",             # mast cell
    "SPINT2", "MKI67", "KRT8",    # epithelial / cancer
    "SOX10", "TP63",              # myoepithelial / basal
    "CD24",                       # B cell + cancer stemness
    "PDCD1"                       # PD-1 (ICI target)
)

p_markers <- sc_seu_subset |>
    BadranSeq::do_FeaturePlot(
        reduction = "umap",
        features = feats,
        order = TRUE,
        ncol = 3
    )

ggsave(
    filename = "../write/figures/02-sc-seu-marker-genes-umap.png",
    plot = p_markers,
    width = 20,
    height = 30,
    bg = "white"
)
```

# Differential Expression

- **`FindAllMarkers`**: one-vs-all Wilcoxon rank-sum on cluster labels — the default Seurat marker detection
- **Cut-offs**: `logfc.threshold = 0.25` and `min.pct = 0.1` are Seurat defaults; tighter thresholds would drop lowly-expressed but lineage-defining markers (e.g. `PROX1` for LECs)
- **Export**: the top 10 markers per cluster (ranked by adjusted p-value) are written to `../write/tables/02-sc-seu-markers.csv` for manual review and as an audit trail of what drove the cluster-to-celltype mapping

```{r}
#| label: dea

# Libraries ----------
library(Seurat)
library(dplyr)
library(readr)

# Processing ----------
markers_df <- sc_seu_subset |>
    FindAllMarkers(
        test.use = "wilcox",
        logfc.threshold = 0.25,
        min.pct = 0.1,
        group.by = "seurat_clusters"
    )

# Outputs ----------
markers_df |>
    group_by(cluster) |>
    slice_head(n = 10) |>
    ungroup() |>
    select(cluster, gene, avg_log2FC, p_val_adj) |>
    arrange(cluster, p_val_adj) |>
    write_csv("../write/tables/02-sc-seu-markers.csv")
```

# Cell Type Annotation

- **Method**: manual `case_when` mapping from `seurat_clusters` to cell type, informed by inspecting the marker UMAPs and the exported marker table
- **Mapping** (for this specific resolution; **re-verify if re-clustered**):
  - Clusters 0, 1, 3, 9 → **cancer_epithelial** — strong `KRT8` / `MKI67` / `CD24`
  - Cluster 2 → **CD4T** — `CD3E` + `CD4` without `CD8A`
  - Cluster 4 → **endothelial** — `RAMP2`+ (no `PROX1`, so blood not lymphatic)
  - Cluster 5 → **myeloid** — `TYROBP`+
  - Cluster 6 → **mesenchymal** — `DCN` / `SFRP2` / `TAGLN`
  - Cluster 7 → **B** — `MS4A1`+ without `JCHAIN`
  - Cluster 8 → **plasmablast** — `JCHAIN`+ (+ residual B markers)
- **Not yet assigned**: mast cells (`CPA3` / `TPSAB1`) did not form their own cluster at this resolution — likely too rare in this sample to separate at `resolution = 0.1`
- **Naming convention**: snake_case for compound labels (`cancer_epithelial`) so they survive downstream metadata joins without needing quoting

```{r}
#| label: annotate

# Libraries ----------
library(Seurat)
library(dplyr)
library(ggplot2)

# Processing ----------
sc_seu_subset$celltype <- case_when(
    sc_seu_subset$seurat_clusters %in% c(0, 1, 3, 9) ~ "cancer_epithelial",
    sc_seu_subset$seurat_clusters == 2               ~ "CD4T",
    sc_seu_subset$seurat_clusters == 4               ~ "endothelial",
    sc_seu_subset$seurat_clusters == 5               ~ "myeloid",
    sc_seu_subset$seurat_clusters == 6               ~ "mesenchymal",
    sc_seu_subset$seurat_clusters == 7               ~ "B",
    sc_seu_subset$seurat_clusters == 8               ~ "plasmablast"
)

# UMAP coloured by celltype — final visual confirmation before checkpoint.
p_umap_celltype <- sc_seu_subset |>
    BadranSeq::do_DimPlot(
        reduction = "umap",
        group.by = "celltype"
    )

ggsave(
    filename = "../write/figures/02-sc-seu-umap-celltype.png",
    plot = p_umap_celltype,
    width = 8,
    height = 6,
    bg = "white"
)
```

# Save Checkpoint

- **`write_rds` (not `saveGiotto`) here**: Seurat objects serialise cleanly through `saveRDS` — the pointer-based-asset problem that makes `saveGiotto` necessary is specific to Giotto's raster images and arrow-backed transcripts
- **`compress = "gz"`**: trades a small save/load time cost for a substantial on-disk saving; the annotated Seurat object is ~1 GB uncompressed

```{r}
#| label: save-checkpoint

# Libraries ----------
library(readr)

# Outputs ----------
write_rds(
    x = sc_seu_subset,
    file = "../checkpoints/02-sc-seu-annotated.rds",
    compress = "gz"
)
```
