---
title: "03 — scRNA-Seq Giotto Object + Annotation Transfer"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    self-contained: true
    title-block-banner: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

# Context

- **Scope**: build a Giotto-native scRNA-Seq object from the same matched h5 used in script `02`, then copy across the cell type labels produced by the Seurat arm
- **Why this script is needed even though Seurat already holds the labels**: the joint integration in script `04` uses `joinGiottoObjects()`, which requires both datasets to be Giotto objects with compatible slots — there is no direct Seurat→Giotto join path
- **Division of labour**:
  - Script `02` (Seurat) — cluster and annotate
  - Script `03` (this one) — rebuild the sc object in Giotto, attach the Seurat labels, save
  - Script `04` (integration) — filter to shared features, `joinGiottoObjects`, Harmony, label transfer onto Xenium cells
- **Upstream**: `../checkpoints/02-sc-seu-annotated.rds` (script 02)
- **Downstream**: `../checkpoints/03-sc-giotto-annotated/` → consumed by `04-integration.qmd`

## Checkpoint Convention

| Stage | File | Description |
|---|---|---|
| Annotated Giotto sc (this script output) | `../checkpoints/03-sc-giotto-annotated/` | Giotto object carrying Seurat-derived `celltype` metadata |

## Dropped approaches (flagged for audit)

- **Giotto-only clustering + annotation** — Leiden partitions did not match published cluster assignments; no clean reference alignment
- **`GPTCelltype` annotation of Giotto marker tables** — GPT-4 produced unstable / unspecific cell type calls, did not outperform manual marker review
- **Consequence**: annotation authority lives in the Seurat arm (script 02); the Giotto sc object is a metadata destination, not an independent analytical path

# Setup

- **Python backend and instructions**: mirrors script `01` — plots route to `../write/figures/`, no on-screen rendering during chunk execution
- **Instructions object is reused**: the same `instrs` configuration is applied to the scRNA-Seq Giotto object for visual consistency with the Xenium arm

```{r}
#| label: setup-giotto

# Libraries ----------
library(Giotto)

# Processing ----------
python_path <- ".venv/bin/python"
set_giotto_python_path(python_path = python_path)

results_folder <- "../write/figures/"

instrs <- createGiottoInstructions(
    save_dir = results_folder,
    save_plot = TRUE,
    show_plot = FALSE,
    return_plot = FALSE,
    python_path = python_path
)
```

# Load scRNA-Seq Data

- **Input**: `../read/GSM7782696_5p_count_filtered_feature_bc_matrix.h5` — the same 10X filtered matrix consumed by script `02`
- **`get10Xmatrix_h5(...)$\`Gene Expression\``**: the 10X h5 can contain multiple feature types (Gene Expression, Antibody Capture, HTO, etc.); explicit selection avoids a subtle mismatch when downstream expects only gene-level counts
- **Minimal construction**: only the expression matrix is loaded — no spatial coordinates, because this is a dissociated scRNA-Seq dataset; Giotto still wraps it as a `giotto` object so the slot structure matches what `joinGiottoObjects()` expects

```{r}
#| label: load-sc

# Libraries ----------
library(Giotto)

# Inputs ----------
expr_data_path <- "../read/GSM7782696_5p_count_filtered_feature_bc_matrix.h5"

# Processing ----------
sc <- createGiottoObject(
    expression = get10Xmatrix_h5(expr_data_path)$`Gene Expression`,
    instructions = instrs
)
```

# Quality Control

- **Metric**: mitochondrial transcript fraction per cell, using `addFeatsPerc` over gene IDs matching `^MT-` (human nuclear mito-prefix convention)
- **Threshold**: `mito < 50%` — deliberately matched to the Seurat arm (script 02) so the two single-cell objects drop the same stressed / dying cells; cross-arm comparability is the priority
- **Expected effect**: ~5,000 cells removed, ~18,000 retained (figures from the Giotto-only exploration)
- **Why two violin plots**: a before/after pair is the simplest possible visual audit that the filter did what we expected — the retained distribution should have `mito < 50` everywhere

```{r}
#| label: qc-mito

# Libraries ----------
library(Giotto)
library(ggplot2)
library(dplyr)
library(data.table)

# Processing ----------
# Identify mitochondrial genes by prefix, then compute per-cell % of reads from them.
fids <- featIDs(sc)
mt_genes <- fids[grep("^MT-.*", fids)]
sc <- addFeatsPerc(
    sc,
    expression_values = "raw",
    feats = mt_genes,
    vector_name = "mito"
)

# Violin — full distribution before filtration.
p_mito_pre <- ggplot(
    spatValues(sc, feats = "mito"),
    aes(x = "cells", y = mito)
) +
    geom_violin() +
    labs(title = "Mitochondrial % — pre-filter", x = NULL, y = "% mitochondrial UMI")

ggsave(
    filename = "../write/figures/03-sc-giotto-mito-pre-filter.png",
    plot = p_mito_pre,
    width = 5,
    height = 6,
    bg = "white"
)

# Count how many cells fall on each side of the cutoff — printed for a permanent
# audit trail in the rendered HTML.
meta <- pDataDT(sc)
cat("Cells with mito > 50 (to be filtered out):",
    meta |> filter(mito > 50) |> nrow(), "\n")
cat("Cells with mito < 50 (to be kept):       ",
    meta |> filter(mito < 50) |> nrow(), "\n")

sc <- subset(sc, subset = mito < 50)

# Violin — after filter. Should show the same shape truncated at 50%.
p_mito_post <- ggplot(
    spatValues(sc, feats = "mito"),
    aes(x = "cells", y = mito)
) +
    geom_violin() +
    labs(title = "Mitochondrial % — post-filter (< 50%)", x = NULL, y = "% mitochondrial UMI")

ggsave(
    filename = "../write/figures/03-sc-giotto-mito-post-filter.png",
    plot = p_mito_post,
    width = 5,
    height = 6,
    bg = "white"
)
```

## Scrublet — currently disabled

- **Intended step**: `doScrubletDetect()` to flag / remove predicted doublets before downstream processing
- **Current status**: crashes R in the uv-managed Python environment — cause traced to an interaction between `scrublet`'s PyTorch dependency chain and reticulate's handling of the Giotto Python backend
- **Decision**: left commented out until a clean fix is found; downstream clustering appears robust enough without it for this sample

```{r}
#| label: scrublet-disabled

# Libraries ----------
library(Giotto)

# Processing ----------
# sc <- doScrubletDetect(sc, return_gobject = TRUE, seed = 0)
```

# Filter Combinations — Threshold Exploration

- **Purpose**: Giotto's `filterCombinations()` sweeps `(expression_threshold, feat_det_in_min_cells, min_det_feats_per_cell)` tuples and plots how many cells / genes survive each — use it to calibrate the actual filter thresholds below
- **Swept values here**: `feat_det_in_min_cells = c(20, 100)` — contrasts a lenient (detected in ≥20 cells) vs. stricter (≥100) gene filter; the lenient setting is chosen for the processing pass since it preserves rare lineage markers

```{r}
#| label: filter-combinations

# Libraries ----------
library(Giotto)

# Processing ----------
filterCombinations(
    sc,
    expression_thresholds = 1,
    feat_det_in_min_cells = c(20, 100),
    save_param = list(
        save_name = "03-sc-giotto-filter-combinations",
        base_width = 7,
        base_height = 5
    )
)
```

# Processing Pipeline

- **Filter thresholds**:
  - `expression_threshold = 1` — a gene is "detected" in a cell if it has ≥1 UMI
  - `feat_det_in_min_cells = 20` — drop genes not seen in ≥20 cells (removes dropout noise while keeping rare lineage markers)
  - `min_det_feats_per_cell = 200` — drop cells expressing <200 genes (stricter than the Xenium panel because scRNA-Seq is transcriptome-wide)
- **`calculateHVF`**: HVF selection is used here (unlike the Xenium arm) because the transcriptome-wide space (~30k genes) is large enough that noise-dominated features will otherwise drown the signal in PCA
- **`runPCA`**: uses the HVF-selected features by default
- **`screePlot(ncp = 50)`**: look further out than the Xenium arm because scRNA-Seq PCs decay more slowly

```{r}
#| label: filter-normalise-pca

# Libraries ----------
library(Giotto)

# Processing ----------
sc <- sc |>
    filterGiotto(
        expression_threshold = 1,
        feat_det_in_min_cells = 20,
        min_det_feats_per_cell = 200
    ) |>
    normalizeGiotto() |>
    calculateHVF() |>
    runPCA()

screePlot(
    sc,
    ncp = 50,
    save_param = list(
        save_name = "03-sc-giotto-scree-plot",
        base_width = 7,
        base_height = 5
    )
)
```

| Parameter | Value | Justification |
|---|---|---|
| `expression_threshold` | 1 | Single-UMI detection standard |
| `feat_det_in_min_cells` | 20 | Lenient; preserves rare markers |
| `min_det_feats_per_cell` | 200 | Standard scRNA-Seq cutoff |
| `ncp` (scree) | 50 | scRNA-Seq PCs decay slowly |

## Dimensional Reduction

- **`dimensions_to_use = seq(30)`**: 30 PCs is the standard scRNA-Seq default — more than the Xenium arm (10) because the transcriptome-wide feature space holds more dimensions of real biology
- **`runUMAP` only (no tSNE)**: UMAP preserves the global structure relevant to the later Harmony integration; tSNE would be purely decorative

```{r}
#| label: umap

# Libraries ----------
library(Giotto)

# Processing ----------
sc <- runUMAP(sc, dimensions_to_use = seq(30))

dimPlot2D(
    sc,
    save_param = list(
        save_name = "03-sc-giotto-umap",
        base_width = 7,
        base_height = 6
    )
)
```

# Clustering

- **`res = 0.5`**: mid-range Leiden resolution; the Giotto clusters are not the annotation source (Seurat is), so the exact partitioning here is less consequential than in script `01` — the goal is just to have a coherent embedding structure to inspect
- **Cluster plot retained** as a diagnostic for the annotation-transfer step: if Giotto clusters and Seurat-derived celltypes disagree sharply, it flags an issue in the transfer join

```{r}
#| label: clustering

# Libraries ----------
library(Giotto)

# Processing ----------
sc <- sc |>
    createNearestNetwork(dimensions_to_use = seq(30)) |>
    doLeidenCluster(res = 0.5)

dimPlot2D(
    sc,
    cell_color = "leiden_clus",
    point_size = 0.6,
    point_border_stroke = 0,
    save_param = list(
        save_name = "03-sc-giotto-umap-leiden-clusters",
        base_width = 8,
        base_height = 6
    )
)
```

# Transferring Seurat Annotations onto the Giotto sc Object

- **Strategy**: cell-ID join — both the Giotto and Seurat objects were built from the same 10X h5, so their cell barcodes are the same strings and can be intersected directly
- **Why `intersect` is still needed**: the two arms apply different filters (`min_det_feats_per_cell = 200` in Giotto; `percent.mt < 50` only in Seurat) — the common cell set is ≤ either individual arm
- **Column-order invariant (`colnames == colnames |> all()`)**: `intersect(giotto_cells, seurat_cells)` preserves the left-hand ordering, so subsetting the Seurat object by that vector yields matching column order — the assertion confirms it
- **`addCellMetadata(by_column = TRUE, column_cell_ID = "cell_ID")`**: joins the annotation table onto Giotto's cell metadata keyed on `cell_ID`; Giotto fills `NA` for any Giotto cell missing from `celltype_df`

```{r}
#| label: annotation-transfer

# Libraries ----------
library(Giotto)
library(Seurat)
library(readr)

# Inputs ----------
sc_seu_annotated <- read_rds("../checkpoints/02-sc-seu-annotated.rds")

# Processing ----------
# Shared cell barcodes between the two arms.
giotto_cells <- sc |> colnames()
seurat_cells <- sc_seu_annotated |> colnames()
common_cells <- intersect(giotto_cells, seurat_cells)

# Base-R subsetting by cell ID (not a tidyverse filter): preserves the exact
# column order from `common_cells`, which is what lets the downstream assertion
# hold. tidyseurat::filter(.cell %in% common_cells) would lose that ordering.
seurat_cells_subset <- sc_seu_annotated[, common_cells]

# Invariant check — fail loudly here rather than silently in the metadata join.
stopifnot(all(colnames(sc) == colnames(seurat_cells_subset)))

# Build the transfer frame. celltype is coerced to character so downstream
# factor ordering does not carry over.
celltype_df <- data.frame(
    cell_ID  = colnames(seurat_cells_subset),
    celltype = as.character(seurat_cells_subset$celltype)
)

sc <- addCellMetadata(
    sc,
    new_metadata   = celltype_df,
    by_column      = TRUE,
    column_cell_ID = "cell_ID"
)

# UMAP coloured by the transferred labels — final visual audit of the join.
dimPlot2D(
    sc,
    cell_color = "celltype",
    point_size = 0.6,
    save_param = list(
        save_name = "03-sc-giotto-umap-celltype",
        base_width = 8,
        base_height = 6
    )
)

# Print metadata to confirm the celltype column landed alongside the existing slots.
print(pDataDT(sc))
```

# Save Checkpoint

- **`saveGiotto`, not `write_rds`**: scRNA-Seq Giotto objects do not carry raster images, but they still carry arrow-backed expression slots; `saveGiotto` is the safe default across all Giotto objects
- **Output path**: `../checkpoints/03-sc-giotto-annotated/` — consumed by `04-integration.qmd`

```{r}
#| label: save-checkpoint

# Libraries ----------
library(Giotto)

# Outputs ----------
saveGiotto(
    gobject = sc,
    dir = "../checkpoints",
    foldername = "03-sc-giotto-annotated",
    method = "saveRDS"
)
```
