---
title: "04 — Joint Integration: scRNA-Seq × Xenium with Harmony"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    self-contained: true
    title-block-banner: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

# Context

- **Scope**: join the processed Xenium object (script `01`) with the annotated Giotto scRNA-Seq object (script `03`) into a single Giotto object, correct the technology-driven batch effect with Harmony, and transfer cell type labels from the scRNA-Seq cells onto the Xenium cells via kNN in the Harmony space
- **Why integrate at all**: Xenium sees ~300 targeted genes; scRNA-Seq sees the transcriptome. Integrating on the shared features lets the Xenium cells inherit annotations learned from a transcriptome-wide view of the same tissue indication
- **Integration anchor**: `vars_use = "list_ID"` — Harmony corrects for the dataset-of-origin variable (the one automatically added by `joinGiottoObjects`); no biological covariates are conditioned on
- **Label transfer**: kNN-based (`labelTransfer`), k = 10, operating on the Harmony embedding — simple and deterministic given the embedding
- **Upstream**:
  - `../checkpoints/01-xenium-gobj/` (script 01)
  - `../checkpoints/03-sc-giotto-annotated/` (script 03)
- **Downstream**: `../checkpoints/04-joined-giotto-harmony-annotated/` — consumed by the advanced spatial-only analyses (`05-advanced-spatial.qmd`)

## Checkpoint Convention

| Stage | File | Description |
|---|---|---|
| Joined object (raw) | `../checkpoints/04-joined-giotto/` | After `joinGiottoObjects` on shared features |
| After Harmony | `../checkpoints/04-joined-giotto-harmony/` | With PCA, pre-/post-Harmony UMAPs |
| After label transfer | `../checkpoints/04-joined-giotto-harmony-annotated/` | Adds `trnsfr_celltype` on Xenium cells |

# Setup

```{r}
#| label: setup-giotto

# Libraries ----------
library(Giotto)

# Processing ----------
python_path <- ".venv/bin/python"
set_giotto_python_path(python_path = python_path)

results_folder <- "../write/figures/"

instrs <- createGiottoInstructions(
    save_dir = results_folder,
    save_plot = TRUE,
    show_plot = FALSE,
    return_plot = FALSE,
    python_path = python_path
)

# Dimensionality budget for the joint workspace — used for PCA, Harmony, UMAP,
# and label transfer so the subspace is consistent across steps.
ndims <- 10
```

# Load Input Checkpoints

- **`loadGiotto(..., reconnect_giottoImage = TRUE)`** on the Xenium object — re-hydrates pointers to the post-Xenium H&E and IF TIFFs in `read/tif_exports/`
- **Images are not strictly needed** for the integration workflow below; reconnecting anyway keeps the loaded object identical in shape to what script 01 produced, which makes it safer to revert to if integration is reworked
- **`python_path` passed to `loadGiotto`**: avoids Giotto defaulting back to its managed conda env on first Python dispatch after reload

```{r}
#| label: load-inputs

# Libraries ----------
library(Giotto)

# Inputs ----------
load_giotto_checkpoint <- function(folder_path, rds_path, reconnect_giottoImage) {
    if (dir.exists(folder_path) && file.exists(file.path(folder_path, "gobject.RDS"))) {
        return(loadGiotto(
            path_to_folder = folder_path,
            reconnect_giottoImage = reconnect_giottoImage,
            python_path = python_path
        ))
    }

    readRDS(rds_path)
}

xenium_gobj <- load_giotto_checkpoint(
    folder_path = "../checkpoints/01-xenium-gobj",
    rds_path = "../checkpoints/01-xenium-gobj.rds",
    reconnect_giottoImage = TRUE
)

sc <- load_giotto_checkpoint(
    folder_path = "../checkpoints/03-sc-giotto-annotated",
    rds_path = "../checkpoints/03-sc-giotto-annotated.rds",
    reconnect_giottoImage = FALSE
)
```

# Feature Intersection and Object Stripping

- **Feature intersection**: Xenium panels cover ~300 curated targets; scRNA-Seq covers the full transcriptome. The join is only defined over their shared features — typically the full Xenium panel (298 features in this sample) because the panel was designed from the transcriptome space
- **Stripping down to core slots**: `joinGiottoObjects` joins on matching slot structures, so we rebuild each object with only the slots needed (`expression`, `spatial_locs`, plus `cell_metadata` on the sc side for the celltype label)
- **`giotto() |> setGiotto(...)`**: empty shell + slot-by-slot injection — a safer pattern than passing the full objects, because the full Xenium object carries images, polygons, and transcript point clouds that the joined object has no business owning
- **`spat_unit = "cell"`** on the Xenium slice: Xenium has multiple spatial units (`cell`, `nucleus`, etc.); only `cell` is used in the join

```{r}
#| label: feature-intersection-and-join

# Libraries ----------
library(Giotto)

# Processing ----------
# Shared features — the join is only defined over these.
ufids <- intersect(featIDs(xenium_gobj), featIDs(sc))
cat("Shared features between sc and Xenium:", length(ufids), "\n")  # expect ~298

# Pull sc cell metadata out before stripping — we want to carry `celltype` through.
sc_meta <- pDataDT(sc)

# Build a minimal sc object carrying only expression + (empty) spatial locs +
# the preserved metadata. Spatial locs are retained even though sc is dissociated
# because joinGiottoObjects expects the slot to be present.
sc_join <- giotto() |>
    setGiotto(sc[[c("expression", "spatial_locs", "cell_metadata"), "raw"]])

sc_join <- addCellMetadata(
    sc_join,
    new_metadata   = sc_meta,
    by_column      = TRUE,
    column_cell_ID = "cell_ID"
)

# Minimal xenium object — only expression + cell-level spatial coordinates.
xen_join <- giotto() |>
    setGiotto(xenium_gobj[[c("expression", "spatial_locs"), "raw", spat_unit = "cell"]])

# Join on shared features. `list_ID` ("sc" / "xen") is auto-added by Giotto
# and becomes the Harmony batch variable further down.
j <- joinGiottoObjects(
    list(sc_join[ufids], xen_join[ufids]),
    gobject_names = c("sc", "xen")
)

# Restore the instructions so plotting in the joined object routes to the same
# figures directory as all other scripts.
instructions(j) <- instrs

# Outputs ----------
saveGiotto(
    gobject = j,
    dir = "../checkpoints",
    foldername = "04-joined-giotto",
    method = "saveRDS"
)
```

# Filter, Normalise, and PCA on the Joined Object

- **Re-filter after join**: filtering is re-applied because the joined object now mixes two technologies with different depth distributions; a unified filter reduces residual bias before Harmony sees the data
  - `expression_threshold = 1` — detection at single-UMI
  - `feat_det_in_min_cells = 1` — very lenient (we already intersected to the Xenium panel, and every gene there is by design a marker)
  - `min_det_feats_per_cell = 10` — low floor; the Xenium cells contribute many fewer unique features per cell than scRNA-Seq
- **Normalisation**: default `normalizeGiotto` (library-size + log) — Pearson residuals and quantile normalisation were tried and did not behave well across the two technologies
- **`runPCA(feats_to_use = NULL)`**: all shared features enter the PCA — no HVF step because 298 panel genes are already lineage-informative by design
- **Pre-Harmony PCA coloured by `list_ID`**: the expected look is two clearly separated clouds — that separation is exactly what Harmony is asked to collapse

```{r}
#| label: filter-normalise-pca-joined

# Libraries ----------
library(Giotto)

# Processing ----------
j <- filterGiotto(
    j,
    expression_threshold = 1,
    feat_det_in_min_cells = 1,
    min_det_feats_per_cell = 10
)

j <- normalizeGiotto(j)

j <- runPCA(j, feats_to_use = NULL)

# PC1/PC2 coloured by dataset — visualises the technology-driven batch effect
# before Harmony correction.
plotPCA(
    j,
    cell_color = "list_ID",
    point_border_stroke = 0,
    point_size = 0.6,
    save_param = list(
        save_name = "04-joined-pca-list-id",
        base_width = 8,
        base_height = 6
    )
)

screePlot(
    j,
    ncp = 50,
    save_param = list(
        save_name = "04-joined-scree-plot",
        base_width = 7,
        base_height = 5
    )
)
```

# Pre-Harmony UMAP

- **Purpose**: snapshot of the embedding *before* batch correction — sc and xenium cells should form mostly disjoint blobs, which is the visual baseline against which Harmony is judged
- **`cell_color_code = c("red", "blue")`**: deliberately high-contrast to emphasise the pre-correction separation

```{r}
#| label: umap-pre-harmony

# Libraries ----------
library(Giotto)

# Processing ----------
j <- runUMAP(j, dimensions_to_use = seq(ndims))

plotUMAP(
    j,
    cell_color = "list_ID",
    point_size = 0.3,
    point_border_stroke = 0,
    cell_color_code = c("red", "blue"),
    save_param = list(
        save_name = "04-joined-umap-pre-harmony-list-id",
        base_width = 8,
        base_height = 6
    )
)
```

# Harmony Batch Correction

- **Why Harmony**: fast, soft-cluster-based integration that works directly on a PCA embedding and produces a corrected low-dimensional space in one call — no reference projection or anchor-finding needed
- **`vars_use = "list_ID"`**: correct only for dataset-of-origin; no biological covariate is conditioned on so that intrinsic population differences survive the correction
- **Post-Harmony UMAP**: the two datasets should now interleave. Residual separation after correction suggests either (a) populations unique to one dataset, or (b) correction under-shot and may need tuning
- **Two post-Harmony views are saved**:
  1. Cells coloured by `list_ID` — the integration quality check
  2. Only sc cells coloured by `celltype`, Xenium shown as low-contrast background — the reference overlay that the label transfer will draw from

```{r}
#| label: harmony

# Libraries ----------
library(Giotto)

# Processing ----------
j <- runGiottoHarmony(
    j,
    vars_use = "list_ID",
    dim_reduction_name = "pca",
    dimensions_to_use = seq(ndims),
    name = "harmony"
)

j <- runUMAP(
    j,
    dim_reduction_to_use = "harmony",
    dim_reduction_name = "harmony",
    name = "umap_harmony",
    dimensions_to_use = seq(ndims)
)

# Integration QC: dataset-of-origin after correction. Cells should intermingle.
plotUMAP(
    j,
    cell_color = "list_ID",
    dim_reduction_name = "umap_harmony",
    cell_color_code = c("red", "blue"),
    point_size = 0.3,
    point_border_stroke = 0,
    save_param = list(
        save_name = "04-joined-umap-harmony-list-id",
        base_width = 8,
        base_height = 6
    )
)

# Palette for the celltype overlays — 10 distinct hues.
cell_colors <- c(
    "turquoise", "azure4", "pink", "tomato",
    "chartreuse3", "lightslateblue", "firebrick4",
    "blue", "magenta", "cornflowerblue"
)

# Reference view: sc cells coloured by celltype, Xenium cells rendered faint.
# This is the visual input to the label transfer — shows the density structure
# the kNN will read from.
plotUMAP(
    j,
    cell_color = "celltype",
    dim_reduction_name = "umap_harmony",
    cell_color_code = cell_colors,
    point_size = 0.5,
    select_cells = spatIDs(j, subset = list_ID == "sc"),
    point_border_stroke = 0,
    other_point_size = 0.3,
    save_param = list(
        save_name = "04-joined-umap-harmony-sc-celltype",
        base_width = 9,
        base_height = 6
    )
)

# Outputs ----------
saveGiotto(
    gobject = j,
    dir = "../checkpoints",
    foldername = "04-joined-giotto-harmony",
    method = "saveRDS"
)
```

# Label Transfer onto Xenium Cells

- **Algorithm**: `labelTransfer` does a kNN lookup in a chosen reduction; each target cell inherits the majority vote of its `k` nearest labelled neighbours plus a posterior probability
- **`source_cell_ids = spatIDs(j, subset = list_ID == "sc")`**: only sc cells contribute labels (by construction, only they carry `celltype`); Xenium cells are the targets
- **`k = 10`**: a conservative neighbourhood — high enough to average over local noise in the kNN, low enough that rare populations (e.g. plasmablasts) don't get smoothed into their more populous neighbours
- **`reduction_method = "harmony"`**: transfer is done in the *integrated* space, not the raw PCA — this is the whole point of running Harmony first
- **Two new columns are added**: `trnsfr_celltype` (the label) and `trnsfr_celltype_prob` (posterior probability) — the probability column is what the advanced spatial analyses use to filter confident transfers from imputed ones

```{r}
#| label: label-transfer

# Libraries ----------
library(Giotto)

# Processing ----------
j <- labelTransfer(
    j,
    source_cell_ids    = spatIDs(j, subset = list_ID == "sc"),
    k                  = 10,
    labels             = "celltype",
    reduction_method   = "harmony",
    reduction_name     = "harmony",
    dimensions_to_use  = seq(ndims)
)

# Visual check — the full joint embedding coloured by the transferred labels.
plotUMAP(
    j,
    cell_color         = "trnsfr_celltype",
    dim_reduction_name = "umap_harmony",
    cell_color_code    = cell_colors,
    point_size         = 0.3,
    point_border_stroke = 0,
    save_param = list(
        save_name = "04-joined-umap-harmony-trnsfr-celltype",
        base_width = 9,
        base_height = 6
    )
)
```

| Parameter | Value | Justification |
|---|---|---|
| `k` | 10 | Averages local kNN noise without erasing rare populations |
| `reduction_method` | `"harmony"` | Transfer in the batch-corrected space, not raw PCA |
| `dimensions_to_use` | `seq(ndims)` (10) | Same subspace used for PCA / UMAP / Harmony |

# Save Final Checkpoint

- **Output**: `../checkpoints/04-joined-giotto-harmony-annotated/` — the downstream input to `05-advanced-spatial.qmd`, which pushes `trnsfr_celltype` back onto the original `xenium_gobj` for spatial-domain analyses
- **Why keep the intermediate checkpoints**: Harmony and label transfer are both deterministic given the inputs but non-trivial to re-run (Harmony is iterative, label transfer is O(n²) in the kNN build); keeping `joined_giotto_harmony/` around lets us re-run only the label-transfer stage with different `k`

```{r}
#| label: save-checkpoint

# Libraries ----------
library(Giotto)

# Outputs ----------
saveGiotto(
    gobject = j,
    dir = "../checkpoints",
    foldername = "04-joined-giotto-harmony-annotated",
    method = "saveRDS"
)
```
